Learning and comparing functional connectomes across subjects

### 2012-12-31 paper.tex
minor edits

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Neuroimage-like layout
%\documentclass[5p]{elsarticle}
% For kindle
\documentclass[1p,12pt]{elsarticle}
\usepackage{geometry}
\geometry{a6paper,hmargin={.2cm,.2cm},vmargin={1cm,1cm}}
% End kindle
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{bm}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{url}
\usepackage[breaklinks=true,letterpaper=true,colorlinks,bookmarks=false]{hyperref}
\usepackage[table]{xcolor}

@@

A first step to move from purely descriptive statistics to modeling with
functional connectivity analysis is to consider a correlation matrix as a
Gaussian graphical model, \emph{i.e.} a well-defined probabilistic model
that describes observed correlations in terms of an independence
structure and conditional relations \cite{lauritzen1996,varoquaux2011}.
In such settings, the inverse covariance graph or the partial
correlations are a measure of influence from one node to another, albeit
undirected. Inferring direction in a Gaussian model is impossible. Linear
structural equation models (SEMs) \cite{mcintosh1994} rely on similar
model that consists in specifying a candidate directed graphical
structure. This structure constraints the covariance matrix of the
signals and can thus be tested on observed data. In fact some forms of
SEMs are known as ``covariance structure models''. Their is thus a strong
formal link between correlation analysis in the framework of graphical
models and SEMs: the former is undirected but fully exploratory, as it
does not require the specification of candidate structure, while the
latter is directory but confirmatory. The link has been exploited to
specify candidate structures for SEMs using partial correlations
\cite{marrelec2007}. More complex models, such as dynamical causal models
(DCMs) \cite{friston2003a} or Granger causality \cite{goebel2003} require
additional hypotheses such as non-linear couplings or time lags.

Most importantly, more complex models can only be used to model
interactions between a smaller number of nodes. This is not only due to a
computational difficulty, but also to fundamental roadblocks in
statistics: the complexity of the model must match the richness of the
data. While injecting prior information can help model estimation, the
more informative this prior, the fragile the inference becomes. The
ongoing debate on the impact of hemodynamic lag on Granger-causality
inference \cite{smith2012a} is an example of such fragility.
While most of the theory underpinning correlation analysis (Gaussian
graphical models) is based on a Gaussian assumption, the core results
are robust to violations of this assumption \cite{ravikumar2011}.

It is tempting to favor more neurobiologically-inspired models that give
descriptions close to our knowledge of the brain basic mechanisms,
however, as George Box famously said, ``all models are wrong; some models
are useful''. Depending on the question and the data at hand, the cursor
should be set between complex models based on a bio-physical description,
and simple phenomenological models such as correlation matrices. In
particular to model interactions between a large number of regions, as in
full-brain analysis, and learn a large \emph{connectome}, simple models
are to be preferred. For more hypothesis-driven studies, such as the
analysis of the mechanism underlying a specific task, more complex models
can be preferred, if rich data is available. Automatic choice of model is
a difficult problem, however, cross-validation (as used in
\cite{varoquaux2010c,craddock2011,strother2006}) is a useful tool. The
central principle of cross-validation is to learn (or fit) a model, and
then test its test predictive power on unseen data. Models too complex
for the data available while fit noise in the data, and thus generalize
poorly. The main benefit of cross-validation is that it is a
non-parametric method which does not rely strongly on modeling
assumption\footnote{This is to be contrasted to Bayesian model
comparison, which will give correct results only if the true generative
model is in the list of models compared. \cite{friston2012} argues that,
based on the Neyman-Pearson lemma, cross-validation is less powerful than
likelihood ratio tests relying on the full dataset. However, it is
important to keep in mind that these approaches only test for
self-consistence, as they require that model likelihood to describe the
data well.}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Neuroimage-like layout
\documentclass[5p]{elsarticle}
% For kindle
%\documentclass[1p,12pt]{elsarticle}
%\usepackage{geometry}
%\geometry{a6paper,hmargin={.2cm,.2cm},vmargin={1cm,1cm}}
% End kindle
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{bm}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{url}
\usepackage[breaklinks=true,letterpaper=true,colorlinks,bookmarks=false]{hyperref}
\usepackage[table]{xcolor}

@@

A first step to move from purely descriptive statistics to modeling with
functional connectivity analysis is to consider a correlation matrix as a
Gaussian graphical model, \emph{i.e.}\ a well-defined probabilistic model
that describes observed correlations in terms of an independence
structure and conditional relations \cite{lauritzen1996,varoquaux2011}.
In such settings, the inverse covariance graph or the partial
correlations are a measure of influence from one node to another, albeit
undirected. Inferring direction in a Gaussian model is impossible. Linear
structural equation models (SEMs) \cite{mcintosh1994} rely on a similar
model that consists in specifying a candidate directed graphical
structure. This structure constraints the covariance matrix of the
signals and can thus be tested on observed data. In fact some forms of
SEMs are known as ``covariance structure models''. There is thus a strong
formal link between correlation analysis in the framework of graphical
models and SEMs: the former is undirected but fully exploratory, as it
does not require the specification of candidate structure, while the
latter is directed but confirmatory. This link has been exploited to
specify candidate structures for SEMs using partial correlations
\cite{marrelec2007}. More complex models, such as dynamical causal models
(DCMs) \cite{friston2003a} or Granger causality \cite{goebel2003} require
additional hypotheses such as non-linear couplings or time lags.

Most importantly, more complex models can only be used to model
interactions between a small number of nodes. This is not only due to a
computational difficulty, but also to fundamental roadblocks in
statistics: the complexity of the model must match the richness of the
data. While injecting prior information can help model estimation, the
more informative this prior the more fragile the inference becomes. The
ongoing debate on the impact of hemodynamic lag on Granger-causality
inference \cite{smith2012a} is an example of such fragility.
Note that although most of the theory underpinning correlation analysis
(Gaussian graphical models) is based on a Gaussian assumption, the core results
are robust to violations of this assumption \cite{ravikumar2011}.

It is tempting to favor more neurobiologically-inspired models that give
descriptions close to our knowledge of the brain basic mechanisms,
however, as George Box famously said, ``all models are wrong; some models
are useful''. Depending on the question and the data at hand, the cursor
should be set between complex models based on a bio-physical description,
and simple phenomenological models such as correlation matrices. In
particular to model interactions between a large number of regions, as in
full-brain analysis, and learn a large \emph{connectome}, simple models
are to be preferred. For more hypothesis-driven studies, such as the
analysis of the mechanism underlying a specific task, more complex models
can be preferred, if rich data is available. Automatic choice of model is
a difficult problem, however, cross-validation (as used in
\cite{varoquaux2010c,craddock2011,strother2006}) is a useful tool. The
central principle of cross-validation is to learn (or fit) a model, and
then test its predictive power on unseen data. Models too complex
for the data available will fit noise in the data, and thus generalize
poorly. The main benefit of cross-validation is that it is a
non-parametric method which does not rely strongly on modeling
assumptions\footnote{This is to be contrasted to Bayesian model
comparison, which will give correct results only if the true generative
model is in the list of models compared. \cite{friston2012} argues that,
based on the Neyman-Pearson lemma, cross-validation is less powerful than
likelihood ratio tests relying on the full dataset. However, it is
important to keep in mind that these approaches only test for
self-consistence, as the Neyman-Pearson lemma requires that 
the model likelihood describes the data well which may not always be the
case\cite{lohmann2012}.}.
